---
title: "Examining Air Quality Trends in the US"
author: "Noah Hansen, Yuhan Dai, Steven Hizmi"
date: "4/19/2021"
output: html_document
---

```{r setup, include=F, message=F}
knitr::opts_chunk$set(echo = F, message = F, warning = F)
library(tidyverse)
library(lubridate)
```


```{r include=F, eval= F}
# used to combine separate csvs into one. Combine the 40 years of daily Aqi data

# library(plyr)
# # from https://statisticsglobe.com/merge-csv-files-in-r
# 
# data_all <- list.files(path = "D:/stat340/data",     # Identify all csv files in folder
#                        pattern = "*.csv", full.names = TRUE) %>%
#   lapply(read_csv) %>%                              # Store all files in list
#   bind_rows                                         # Combine data sets into one data set
# 
# write.csv(data_all, file = "D:/stat340/data/aqi_1980_2019", sep=",")
```


```{r}
# read in the daily aqi data from 1980 to 2019

aqi_dat = read_csv("aqi_1980_2019.csv")
aqi_dat = aqi_dat %>% select(-X1, 
                     cbsa = CBSA,
                     date = Date,
                     aqi = AQI,
                     category = Category,
                     cbsa_code = `CBSA Code`, 
                     param = `Defining Parameter`, 
                     site = `Defining Site`,
                     num_sites = `Number of Sites Reporting`)

aqi_dat = aqi_dat %>% mutate(year = year(date), month = month(date), wday = wday(date, label=TRUE))


census_cbsa = read_csv("census_2010.csv") # combine population data

census_cbsa = census_cbsa %>% 
  select(cbsa_code = CBSA, pop = CENSUS2010POP) %>% 
  group_by(cbsa_code) %>% 
  summarise(pop = sum(pop)) # get pop of whole cbsa

aqi_dat = aqi_dat %>% left_join(census_cbsa)
```


```{r include=F, eval = F}
# This table shows the average and maximum gaps in reporting each criteria pollutant for each CBSA. Maybe we would want to exclude some of the pollutants that are not measured very often per CBSA code. We can repeat this analysis for different time periods instead of the whole 40 years, as some stop reporting certain values for years at a time. This would be pretty obvious in the graphs though and can be addressed as it comes up.

time_diff <- aqi_dat %>% 
  group_by(cbsa_code,param) %>% 
  arrange(date) %>% 
  # lag() gets the value from the previous row. 
  # Then extract digits from output of the difference, which is a sentence for some reason
  summarise(date_diff = as.integer(str_extract(date-lag(date),"\\d*"))) %>% 
  summarise(diff_avg = mean(date_diff, na.rm = T), diff_max = max(date_diff, na.rm=T))

time_diff
```


```{r include = F}
# Unclear why this way does not work for date difference

# how to take the differce in dates
# x1 <- as.Date("2019-02-07")
# x2 <- as.Date("2018-02-07")
# x1 +1
# abs(julian(x1,x2)[1])
# 
# julian_new = function(date, origin){
#   return(tryCatch(expr = julian(date,origin), error = function(e){print(e); return(NA)}))
# }

# aqi_dat %>% mutate(date_diff = abs(julian_new(date,lag(date))[1])) 
```


# Introduction


# Data

[census data](https://www.census.gov/data/tables/time-series/demo/popest/2010s-total-metro-and-micro-statistical-areas.html)
[Daily and Annual Ozone data](https://aqs.epa.gov/aqsweb/airdata/download_files.html)
[US Vehicle Miles](https://www.bts.gov/content/us-vehicle-miles )

TODO- Explain relevant pollutants.

- CO: Carbon Monoxide forms from burning fuel. 

- NO2: Nitrogen Dioxide forms from burning fuel. It can cause irritation of the throat, nose, and eyes, and decrease lung function when inhaled. 

- Ozone: Ozone, known for creating smog, forms from the sunlight reacting with gases, such as nitrogen oxides. Ozone causes lung damage when inhaled in relatively low amounts.

- PM10: Fine particulate matter (diamater $<=$ 10 micrometers) can come from combustion, but also includes dust from construction, agricuture, and even pollen. PM10 deposits on the surfaces of the larger airways of the upper region of the lung. It can worsen existing respitory diseases.

- PM2.5: Particulate matter (diamater $<=$ 2.5 micrometers) forms from burning fuel. It is more likely than PM10 to cause damage by depositing deeper in the lung. This is usually the particle that you would associate with air pollution as it causes the most problems, mostly due to its prevalance. 

- SO2: Sulfure Dioxide

-- Cleaning the Daily and Annual AQI datasets

- Combined last 6 years of annual data using rbind(), and filtered the data for the 3 counties in Wisconsin that we focused on

- Used separate data frames for 2019 and 2020 daily AQI data

-- Cleaning the US Vehicle Miles data

- Adding row names to the data, reshaping the dataset by using transpose and filtered the data for 4 variables w/o NA due to the data set is ragged on both edge.

# Methods

## Steve

-- Used regression with geom_smooth when analyzing last 6 years of annual AQI data. Chose this method to see the relationship between 2020 data and the years that preceded that, as well as checking similarities within those previous years.

-- Used a scatterplot for the daily data, so we can see any clusters of datapoints and how that compared to clusters of data during Covid in the first half of 2020.

## Yuhan
-- Used linear regression on CO and PM2.5 pollutants with miles being the explanatory variable to discover the relationship between them. 

-- Used lineplot to depict the trend of AQI level for each pollutant for the entire US for the last 4 decades, and used scatter plot for the CO, PM2.5 and miles, and then added linear regression line to fit the scatter plot.

# Results

## Noah

First, we examine air quality changes from 1980 to 2019 for The US as a whole. Since there are 6 different pollutants, and many more testing sites, we take the average of the AQI values for each day. This is likely not a perfect heuristic, because a dangerous value for one pollutant should not be diminished by a safe value of another. However, if overall pollution levels change significantly it should be reflected in the mean.


```{r}
aqi_summary = aqi_dat %>% group_by(date) %>% 
  summarise(aqi = mean(aqi)) 
 
aqi_summary %>% ggplot(aes(x = date, y = aqi)) +
  geom_point(alpha = .2) +
  geom_smooth(color = "red", method = "lm") +  # se = F
  labs(title = "Average AQI over Time", x = "Year", y = "AQI")
```

```{r include = F}
lm.fit = lm(aqi ~ date, data = aqi_dat)
summary(lm.fit)
coef(lm.fit)[[2]]*365*40


head(aqi_dat)

pred = tibble(
  date = c(ymd("2060-01-01"))
)

predict.lm(lm.fit, pred)
```
Pollution data is inherently noisy, as levels can fluctuate widely from day to day. The wind could blow a high level of pollution from one area to another, for example, or there might be a forest fire every few years that increases levels. However, the downward trend estimate of about -.55 per year (P < .001) has a small standard error. Even noisy, high-variability data can have a significant trend. While -.55 is a small change in one year, over 40 years it has led to about a 22 point decrease in the aqi levels. If the trend continues, by 2060 the average will be around 17. It is also promising how the variability decreases after 2012 as this suggests less extreme AQI values. 

```{r}
aqi_dat %>% group_by(date, param) %>% 
  summarise(aqi = mean(aqi)) %>% 
  ggplot(aes(x = date, y = aqi)) +
  geom_point(alpha = .15) +
  geom_smooth(color = "red", method = "lm") +  # se = F
  facet_wrap(~ param) +
  ylim(0,125) +
  labs(title = "Average AQI over Time, by Pollutant", x = "Year", y = "AQI")
```

Now lets relate pollution levels to the population of cities.


```{r}
top_10 = aqi_dat %>% group_by(cbsa_code) %>% summarise(pop = pop[1], cbsa = cbsa[1]) %>% slice_max(n=6, order_by = pop) %>% select(cbsa_code, cbsa)

# top_10 %>% group_by(date, param) %>% 
#   summarise(aqi = mean(aqi)) %>% 
#   ggplot(aes(x = date, y = aqi)) +
#   geom_point(alpha = .15) +
#   geom_smooth(color = "red", method = "lm") +  # se = F
#   facet_wrap(~ param) +
#   ylim(0,125) +
#   labs(title = "Average AQI over Time, by Pollutant", x = "Year", y = "AQI")

top_10
```

```{r}
aqi_dat %>% 
  filter(year == 2019, param == "Ozone") %>% 
  group_by(cbsa_code) %>% 
  summarise(aqi = mean(aqi), pop) %>% 
  ggplot(aes(x = pop, y = aqi)) + 
  geom_point(alpha = .2) +
  geom_smooth(method = "lm")
```

Here we are able to look at the maximum reported values for each pollutant for every day in the New York area.
```{r}
# for New York-Newark-Jersey City

aqi_dat %>% filter(cbsa_code == 35620) %>% 
  group_by(date, param) %>% 
  summarise(aqi = max(aqi), cbsa) %>% 
  ggplot(aes(x = date, y = aqi)) +
  geom_point(alpha = .2) +
  geom_smooth(method = "lm") +
  facet_wrap(~ param) 
```


## Han

The following section groups the data by year and pollution type, then summarized the data by taking the mean of the AQI. 
The graph depicts the change of air quality data over the last 40 years. The horizontal line indicates the Good Level for AQI, below the line (50) means the air quality is satisfactory.
The details regarding the levels of concern for AQI can be found here: https://www.airnow.gov/aqi/aqi-basics/

```{r}
yr_pol = aqi_dat %>% group_by(year, param) %>% summarise(avg_aqi = mean(aqi))

ggplot(yr_pol, aes(x = year, y = avg_aqi, color = param)) +
  geom_line(size = 1) +
  geom_hline(yintercept = 50) +
  labs(title = "US Air Quality Data by Pollution Type")
```

As shown above, US pollution overall is improving since the AQI is decreasing for most of the pollution type. However, there are drastic increases for both CO and PM2.5. 

In order to understand such change, I decided to look into the main source of CO and PM2.5 fossil fuels. For this part, US vehicle miles are used to represent fossil fuel usage. The following section is to discover the trend of US vehicle miles from 1960 to 2019.  
The data used in this section is US vehicle miles.

```{r, include=F}
vehicle_miles = read_csv("US_vehicle_miles.csv")
#clean up data
names(vehicle_miles) <- as.character(unlist(vehicle_miles[1,]))
vehicle_miles = vehicle_miles[-1,]

vehicle_use <- vehicle_miles %>% filter(Year == "Air carrier, domestic, all services" | Year == "Highway, total" | Year == "Transite, total" | Year == "Total train-miles")
# want the sum of miles for each category
t_vehicle <- as.data.frame(t(vehicle_use))
names(t_vehicle) <- as.character(unlist(t_vehicle[1, ]))
rowname = rownames(t_vehicle)
t_vehicle = t_vehicle[-1,]

for (i in 1:nrow(t_vehicle)){
  for (j in 1:ncol(t_vehicle)){
      t_vehicle[i, j] = gsub(",","",t_vehicle[i, j])
  }
}
#convert to numeric
for (j in 1:ncol(t_vehicle)){
  t_vehicle[,j] = as.numeric(t_vehicle[,j])
}

t_vehicle <- t_vehicle %>% mutate(sum = rowSums(.), year = rowname[-1])
```

```{r}
ggplot(t_vehicle, aes(x = year, y = sum)) +
  scale_y_continuous(trans='log10') +
  geom_point() +
  geom_line(aes(group = 1)) +
  labs(title = "U.S. Vehicle-Miles") +
  xlab("Year") +
  ylab("Miles (Millions)") +
  theme(axis.text.x = element_text(angle = 60, hjust = 1))
  
```

As shown above, the usage of vehicles in the US is significantly increasing over the past few decades. Due to the source of the data, between year 1960 and 1990, the discrepancy is 5 years instead of 1, which results in a steeper slope than from 1990 to 2019.  
  
  
The following section is to perform a linear regression on CO and PM2.5 level with US vehicle miles to see if there's a significant relation between them. 

```{r, include=F}
#CO and PM2.5 data
poll2 <- yr_pol %>% filter(param == "PM2.5" | param == "CO") %>%
  pivot_wider(names_from = param, values_from = avg_aqi) %>%
  filter(year >= 1990)

temp_miles <- t_vehicle %>% 
  select(sum, year) %>% 
  filter(year >= 1990) %>% 
  rename(miles = sum)

mile_poll <- merge(poll2, temp_miles, by = "year")

lmCO = lm(CO ~ miles, data = mile_poll)
summary(lmCO)

lmPM = lm(PM2.5 ~ miles, data = mile_poll)
summary(lmPM)
```

According to the linear regressions above, the p values of miles for each pollutant are significant enough (1.73e-09 and 0.0010) which indicates that there's a correlation between US vehicle usage and pollutant level for PM2.5 and CO. However, the slope for CO is negative. To better illustrate, here is the plot of the lm.

```{r}
ggplot(data = mile_poll, aes(x = miles, y = CO))+
  geom_point() +
  geom_smooth(method = 'lm', level=0.95)

ggplot(data = mile_poll, aes(x = miles, y = PM2.5))+
  geom_point() +
  geom_smooth(method = 'lm', level=0.95)
```
The plot may seem to be contradictory to the initial assumption which is fossil fuel use is positively related to CO and PM2.5 emission. However, other factors will contribute to this trend. For instance, the improved combustion efficiency of the engine may lead to less CO emission. In addition, since vehicle emission is not the only source of CO emission, the increase percentage of clean energy may also leads to the decrease of CO's aqi level. 

### Steve

```{r, echo=FALSE}
county_2015 <- read.csv("annual_aqi_by_county_2015.csv")
county_2016 <- read.csv("annual_aqi_by_county_2016.csv")
county_2017 <- read.csv("annual_aqi_by_county_2017.csv")
county_2018 <- read.csv("annual_aqi_by_county_2018.csv")
county_2019 <- read.csv("annual_aqi_by_county_2019.csv")
county_2020 <- read.csv("annual_aqi_by_county_2020.csv")

daily_2019 <- read.csv("daily_44201_2019.csv")
daily_2020 <- read.csv("daily_44201_2020.csv")

counties <- rbind(county_2015,county_2016, county_2017, county_2018, county_2019, county_2020) 

```

**With the lockdowns during the Covid pandemic, it gave us an opportunity to see what happened to air pollution with less human activity in three Wisconsin counties: Dane, Milwaukee, and Waukesha counties. So we have collected county data over the last 6 years to compare with data during the lockdowns. We also collected daily data from 2020, from January to June (as that is the most updated data out there thus far).**

```{r}
wisconsin_counties <- counties %>% filter(County == "Dane" | County == "Milwaukee" | County == "Waukesha")
ggplot(wisconsin_counties, aes(Year, Days.Ozone, color = County)) +
  geom_smooth() +
  ylab("Ozone Action Days") +
  ggtitle("Ozone levels in WI over last 6 years")
  
```

**According to the U.S. Air Quality Index (AQI), Ozone levels at or below 50 represents good air quality whereas Ozone levels at or above 300 represents hazardous air quality. Ozone Action Days are days that are declared to be of caution to citizens, as they should limit their time outdoors. As we can see, ozone action days from 2015 to 2019 stay around 125, with Milwaukee county being higher on average. But in 2020, we see a significant trend downwards. In March of 2020, the Governor of Wisconsin issued a stay at home order, and the closure of all non essential business due to the Covid outbreak. We can clearly see that the decrease in human activity led to less days of harmful Ozone levels in these three Wisconsin counties. To confirm this conclusion, we are going to look at the daily data of these three counties in the first half of 2020 and see if these changes coincide with the stay at home orders. **

```{r}
require(lubridate)
wisconsin_daily <- daily_2020 %>% filter(County.Name == "Dane" | County.Name == "Milwaukee" | County.Name == "Waukesha" ) %>% mutate(Date.Local = month(Date.Local))


ggplot(wisconsin_daily, aes(Date.Local, AQI, color= County.Name)) +
  geom_point() +
  xlab("Month") +
  ggtitle("Ozone levels first 6 months of 2020")

```

**Considering that the first lockdown in Wisconsin was March, the ozone levels don't go beyond 50, which is deemed safe. The ozone levels don't start increasing until May and June, which is to be expected as Ozone levels increase during summer. The question is: Did Covid have an affect on the Ozone levels or was it increased due to it being Summer? To see which is true, we decided to look at the daily data from 2019, with no stay at home orders to see how they compare.**

```{r}
wisconsin_daily_19 <- daily_2019 %>% filter(County.Name == "Dane" | County.Name == "Milwaukee" | County.Name == "Waukesha" ) %>% mutate(Date.Local = month(Date.Local))


ggplot(wisconsin_daily_19, aes(Date.Local, AQI, color = County.Name)) +
  geom_point() +
  xlab("Month") +
  ggtitle("Ozone levels daily in 2019")
```

**Looking at this graph, the Ozone levels are above 50 by mid-February. Compared to 2020, the 2019 Ozone levels increased far more quickly than in 2020. The first case of Covid was reported in February in Wisconsin. So it is possible many people began to stay at home before the stay at home order, which could be an explanation in the disparity between the early months of February in 2019 and 2020. We think without a doubt that Covid had an impact in these Wisconsin counties, especially Milwaukee as it is more industrious than Waukesha and Madison. With data, we have seen the impact that human activity has on air pollution. It can already be seen that stopping human activity for a few months made the Ozone levels stay under 50 far longer than with regular human activity. We suggest that this is enough evidence that we can really make a change for the better, and that lowering pollution levels should remain a priority.**

# Conclusion

Over time, pollution levels have improved in the US for metro and micropolitan areas. Higher density areas, such as New York City, has worse pollution levels and health problems related to pollution. We suggest to continue lowering pollution levels as average pollution levels since 1980 has been on a downward trend, as seen in our first figure. Breaking down the pollution levels by pollutant as seen in our second figure, Carbon Dioxide has been decreasing over time significantly. Its important to reduce as much CO2 as possible, as its a toxic gas to humans, as well as a catalyst to global warming that has bad effects on Earth. Considering that Carbon emissions from cars is significantly higher per year than what it was in 1980, seeing that CO2 pollutants is decreasing is a positive sign and that we should continue to keep lowering air pollution as a priority. An argument against pollutant affects on global warming is that its a part of Earths history, as we have gone through many cycles of warming and cooling periods and that humans are not affecting it. With the Covid pandemic, it gave us a unique opportunity to look at what happens to air pollution levels with less human activity to test if humans have an impact. Looking at the figure titled "Ozone levels in WI over last 6 years" we can see that Ozone levels stay relatively consistent from 2015-2019. In 2020, there is a significant decrease in Ozone levels. Considering the 2020 data is only for the 6 months of 2020, suggests that human activity has a huge impact on our environment especially with a change in levels in just a year.

## describe limitations here

Although most of the pollutants AQI level are steadily decreasing, CO has been fluctuating frequently in the last decade. With the help of regression analysis, we can relatively confident to rule out vehicle emission as a main contributor. However, the main reason that causes the rapid increase of CO emission is yet to be studied. If we take a closer look at the time period in which CO's AQL level is rapidly increasing, we can maybe connect the time frame with real worlds events that are related to carbon emission to further investigate the connection.


# Appendix

If you would like to see the code, it is [hosted on GitHub](https://github.com/noahansen/air-pollution-analysis).


